---
title: "Coronaviruses cases analyzer"
author: "Rick Lee"
output:
  html_document:
    df_print: paged
---

**Coronavirus data analyzer**
Performs various analyses on countries (for US, can also specify state) of covid-19 cases according to user input. The resource can be found in this link, which belongs to Johns Hopkins University: 

https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_daily_reports  


Summary of important variables: 

raw: raw dataset
main, main1: temporarily stores dataset for data cleaning
mainsub: cleaned dataset, filtered to contain only countries of user input

countries: list of available countries for analysis
country: vector containing countries of user input (selected from 'countries')
state: vector containing states if the countries is US (for each element in 'country', if the country is US, the state is stored, else, '.' is stored)

comparison.levels: column added to 'mainsub' to perform comparisons. States (if US) and other countries are assigned in this same column to allow for direct comparisons
Mortality: column added to 'mainsub' as deaths/confirmed cases
Mortalitypercent: column added to 'mainsub' as Mortality*100


Things to finalize: 
- add country names to rows in describeBy
- get describeBy to show more than 2 sig figs



```{r}
library(psych)
```



```{r}
rm(list = ls())

#the file from the link in the description. The sample code takes cases up to December 10th 2020, but this section can be altered and the file downloaded using the link for any other day. 
raw <- read.csv("Corona-12-10-2020.csv", sep = ",", header = TRUE, fill = TRUE, quote = "")


#list of available countries
countries <- c("Australia","Brazil","Canada","Chile","China","Colombia","France","Germany","India","Italy","Japan","Korea","Mexico","Netherlands","Pakistan", "Peru","Russia","Spain","Ukraine","United Kingdom", "US")

#data cleaning: 
raw <- raw[,c(2:4,8:11)]

#eliminating an anomaly with a mortality rate of over 70% - it messes up the box plots
raw <- raw[-6006,]


#more data cleaning
count <- c()
for (x in 1:length(countries)){
    
  for (y in 1:nrow(raw)){
    if (countries[x] == raw$Country_Region[y]){
    
    count <- c(count, y)
    }
  }
}

main <- raw[count,]


for (x in c(4:7)){
main[,x] <- sapply(main[,x], as.numeric)
}
main

main1 <- subset(main, is.na(main[1:nrow(main),4]) == F & is.na(main[1:nrow(main),5]) == F & is.na(main[1:nrow(main),6]) == F & is.na(main[1:nrow(main),7]) == F)


#the list of countries did not include Korea. So I included it (date: 8th January 2021)

koreadata <- data.frame(Admin2 = rep("", 17),Province_State = c('Seoul', 'Busan', 'Daegu', 'Incheon', 'Gwangju', 'Daejeon', 'Ulsan', 'Sejong', 'Gyeonggi', 'Gangwon', 'Chungbuk', 'Chungnam', 'Jeonbuk', 'Jeonnam', 'Gyeongbuk', 'Gyeongnam', 'Jeju'), Country_Region = rep("Korea", 17), Confirmed = c(23459, 2570, 8250, 3679, 1511, 1085, 915, 187, 19063, 1611, 1513, 1982, 1013, 713, 2874, 1914, 518), Deaths = c(301 , 88, 207, 46, 16, 13, 36, 1, 414, 30, 53, 33, 42, 6, 64, 7, 0), Recovered = c(18626, 2057, 7880, 3154, 1305, 913, 804, 159, 15515, 1369, 1218, 1772, 884, 575, 2610, 1649, 488), Active = c(4532, 425 , 163, 479, 190, 159, 75, 27, 3134, 212, 242, 177, 87, 132, 200, 258, 30))

main1 <- rbind(main1, koreadata)

```




```{r}

#taking country (and possibly state) input from user

cat("Welcome to the Coronavirus cases analyzer. \nThe list of available countries for analysis are as follows: \n
Australia, Brazil, Canada, Chile, China, Colombia, France, Germany, India, Italy, Japan, Korea, Mexico, Netherlands, Pakistan, Peru, Russia, Spain, Ukraine, United Kingdom, US \n
Enter any number of countries to compare them. \nIf any of the countries are the US, you can also specify the state. \n")


#taking input for number of countries to be compared: 
repeat{
countrynumber <- readline("How many countries do you want to compare? ")

  if(countrynumber >= 0){
    break
    
  } else{
    cat("Please enter a positive integer! ")
  }
}




#taking input for n number of countries, where n is the previously input number of countries: 
country <- c()

for (x in 1:countrynumber){
country[x] <- readline("Enter any country : ")

  if (country[x] == "UK" | country[x] == "uk"){
    country[x] <- "United Kingdom"
  }

country[x] <- toupper(country[x])
}

countries <- toupper(countries)


#creating dataframe for countries: goes through each row to select the countries that match user input
mainsub <- data.frame()

for (x in 1:length(unique(country))){
y = 0

  while (y <= length(countries)){
    y <- y + 1
    
    if (y == length(countries) + 1){
      
      y <- y + 10000
      stop("Looks like country ",x ," is not a valid country. Did you make sure to select from the list of countries?")
    }
  
    
    if (unique(country)[x] == countries[y]){
      
      mainsub <- rbind(mainsub, subset(main1, toupper(main1$Country_Region) == countries[y] & main1$Province_State != "Unknown"))
      
      y <- y + 10000
      
    } 
  
  }

}

```



```{r}
#processing state input from user if country is US

#note: only runs if any of the inputs are the US
state <- c()
rowstoerase <- c()

for (x in 1:length(country)){
  
  if (country[x] == "US"){
    state[x] <- readline("Enter any state in the US (if unspecified please enter 'no'): ")
    state[x] <- toupper(state[x])
  }else{
    state[x] <- "."
  }

  
  if (country[x] == "US"){
    #cleaning data to get rid of N/As
    for (z in 1:nrow(mainsub)){
      if (mainsub$Country_Region[z] == "US" && is.na(mainsub$Province_State[z]) == TRUE){
        
          mainsub <- mainsub[-z,]
        
    }
  }
  
  
  #erase all rows of US apart from the specified state
  y = 1
    repeat{
      
      #except when state wasn't specified (then the whole of US needs to be included)
      if (state[x] == "NO"){
        break
      }
      
      if (y > nrow(mainsub)){
        break
      }
      
      if (mainsub$Country_Region[y] == "US" && state[x] != toupper(mainsub$Province_State[y])){
        if (y %in% rowstoerase == FALSE){
          
          #variable for rows to erase (rows of US other than the specified state)
          rowstoerase <- c(rowstoerase, y)
        }
      }
      
      y <- y + 1
    }
  }
} 


#in case there were two inputs for US, each with a different state, in order for both the states to not be erased, their rows must be deducted from rowstoerase
for (x in 1:length(country)){
  
  for (y in 1:nrow(mainsub)){
    
    if (mainsub$Country_Region[y] == "US" && state[x] == toupper(mainsub$Province_State[y])){
          if (y %in% rowstoerase == TRUE){
            rowstoerase <- rowstoerase[rowstoerase != y]
          }
    }
    
  }
}

```



```{r}

#in case any of the inputs were US WITHOUT states being specified, then all rows need to be kept 
include.allstates <- FALSE


if("NO" %in% state == TRUE){
  include.allstates <- TRUE
}
  


if('US' %in% country == TRUE && include.allstates == FALSE){
      mainsub <- mainsub[-rowstoerase,]
}

```


```{r}

#make a new column to perform comparisons against
#states (if US) and other countries are assigned in this same column for direct comparisons against each other

comparison.levels <- c()


for (x in 1:nrow(mainsub)){
  
  if (mainsub$Country_Region[x] == 'US' && toupper(mainsub$Province_State[x]) %in% state == TRUE){
    
    #assigning state names to the new column if a US state. 
    comparison.levels <- c(comparison.levels, mainsub$Province_State[x])
    
  } else {
    
    #assigning country names to the new column if not a US state. The thing is, if US without state being specified AND US with a specific state were both inputted, the columns here would overlap. 
    comparison.levels <- c(comparison.levels, mainsub$Country_Region[x])
  }
}
mainsub <- cbind(mainsub, comparison.levels)


# ^ to fix the above problem: if US was one of the options with state not specified, add all rows of US to the end: 

#note: only performed if US with state not specified is an input
if ("NO" %in% state == TRUE){
  
  #creating subset a that would later be rbinded to mainsub: 
  a <- subset(mainsub, mainsub$Country_Region == 'US')
  
  for (x in 1:nrow(a)){
  a$comparison.levels[x] <- 'US'
  
  }
  
  rowstoerase <- c()
  
  for (x in 1:nrow(mainsub)){
    
    #erase rows from original column with comparison column as US 
    #for states that were specified, the comparison column would have the state, not US. 
    #So only US rows that are NOT part of specified states are erased
    
    #since all the US rows were previously stored under subset a, they can be added at the end later. 
    if (mainsub$comparison.levels[x] == 'US'){
      rowstoerase <- c(rowstoerase, x)
    
      }
    
  }

  #erasing US non-specified rows and adding subset a: 
  mainsub <- mainsub[-rowstoerase,]
  
  mainsub <- rbind(mainsub, a)  
  
}


```
```{r}
#input command for function

repeat{
operations <- readline("Enter 1 to display summary statistics, \n2 to perform significance tests, \n3 to generate box-plots, \n4 to display regions with highest cases within each country/state, \nor 5 to do all.\n")

if (operations >= 1 && operations <= 5){
  break
  
} else{
    cat("Please enter a number between 1 and 5! ")
  }
}

```



```{r}
#adding mortality columns
Mortality <- c()


for (x in 1:nrow(mainsub)){
  if (mainsub$Confirmed[x] > 0){
  Mortality <- c(Mortality, mainsub$Deaths[x] / mainsub$Confirmed[x])
  
  } else{
    Mortality <- c(Mortality, 0)
  }
}
Mortalitypercent <- Mortality * 100

mainsub <- cbind(mainsub, Mortality)
mainsub <- cbind(mainsub, Mortalitypercent)

```



```{r}
#function for obtaining the region with highest reported cases

if (operations == 4 | 5){
  
  
    highestregions <- c()
    
    #set beginning point (to the respective country/state)
    for(x in 1:length(unique(mainsub$comparison.levels))){
      a <- 1
      repeat{
        
        if (mainsub$comparison.levels[a] == unique(mainsub$comparison.levels)[x]){
          
          highest_count <- a
          break
        }
        a <- a + 1
      }
        
    
    #update count as highest region
    for(y in 1:nrow(mainsub)){
      if (mainsub$comparison.levels[y] == unique(mainsub$comparison.levels)[x]){
            
        if (mainsub$Confirmed[y] > mainsub$Confirmed[highest_count]){
            
              
        highest_count <- y
            
        }
            
      }
    }
      
    #store vector for outputting highest region  
    if(toupper(unique(mainsub$comparison.levels)[x]) %in% state == TRUE){
      highestregions <- c(highestregions, mainsub[highest_count,1])
      
      } else{
      highestregions <- c(highestregions, mainsub[highest_count,2])
      
      }
    }
  
}

```


```{r}
#function for summary stats using describeBy (it's correcting to 2 sig figs even when I specified the digits to 6??)
#also I have to add names to the rows
if (operations == 1 | operations == 5){

  uniqueregion <- data.frame()
  summarystats <- data.frame()
  
  for(x in 1:length(unique(mainsub$comparison.levels))){
    uniqueregion <- data.frame()
    
    for(y in 1:nrow(mainsub)){
        
        if (mainsub$comparison.levels[y] == unique(mainsub$comparison.levels)[x]){
          uniqueregion <- rbind(uniqueregion, mainsub[y,])
          
        }
        
    }
      summarystats <- rbind(summarystats, describeBy(uniqueregion$Mortality, digits = 6))
  }
  summarystats
}

#function for significance testing
if (operations == 2 | operations == 5){
  cat("\n\nThe following displays significance testing across the countries/states: \n\n")
  
  if (length(unique(mainsub$comparison.levels)) <= 2){
    t.test(mainsub$Mortality ~ mainsub$comparison.levels)
    
  } else{
    
    fit <- aov(mainsub$Mortality ~ mainsub$comparison.levels)
    summary(fit)
    TukeyHSD(fit)
    
  }
  
}

#function for boxplots
if (operations == 3 | operations == 5){
  
  boxplot(Mortalitypercent ~ comparison.levels, data = mainsub, ylab = 'Mortality Rate (%)', xlab = 'Countries', main = 'Mortality Rates for Each Country', col = 'green')
  
}

#outputting regions with highest cases 
if (operations == 4 | operations == 5){
  cat("\nThe following displays the regions with highest cases: \n")
  
for (x in 1:length(unique(mainsub$comparison.levels))){
    cat("\nThe region with highest reported cases in ", 
        
        
          unique(mainsub$comparison.levels)[x], 
        
        
        " is ", highestregions[x], ". ")
  }
}


```





